P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2024-07-04 sage_latex_template/manuscript.tex
Adjust width of figures
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

When computing forces for DEM simulations that contain different shapes, different calculation formulas are needed for some of the contact elements such as the penetration depth, contact point and distance.
From a computational point of view, this introduces branching in the inner-most loop where each path leads to different arithmetic instructions be executed and therefore inhibits SIMD vectorization.
Algorithm \autoref{alg:branching} displays an example of such issue for a kernel with spheres and half-spaces.
Note that when traversing the neighbors, their shapes are checked to determine which code path is taken during the interaction, hence the code is not SIMD-friendly.

@@

To provide a simpler way to achieve vectorized code for such DEM kernels, we partition particles by shape in our acceleration data structures (Linked Cells and Verlet Lists), which allows distinguishing them during the force computation.
After, we use code generation to apply a loop-splitting in the generated code for every pair of shapes, these resulting loops should be then easier to be vectorized since no branches are needed to calculate different contact elements for the particles being computed.
Algorithm \autoref{alg:no_branching} displays the same kernel as Algorithm \autoref{alg:branching} without the branching issue.
Note that the outer-most loop can also be split into two parts as well, which leads to two different kernels for each pair of shapes.
For GPUs, this can be beneficial since different threads will be used for different pair of shapes (parallelism is done at the level of the outer-most loop) and no threads will be disabled due to traversing a different code path.

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.75]
    \pgfplotsset{
        ybar stacked, ymin=0, ymax=32, xmin=0.5, xmax=2.5, xtick=data,
        xtick={1,...,2},
        xticklabels={Ice Lake, Milan},
        xticklabel style={yshift=-10pt},
        ylabel={time to solution (s)}, ylabel style={yshift=-1ex},
        legend cell align={left},
        /pgf/bar width=12pt,% bar width
        scatter/position=absolute,
        node near coords style={
            font=\footnotesize,
            at={(axis cs:\pgfkeysvalueof{/data point/x},\pgfkeysvalueof{/pgfplots/ymin})},
            anchor=north,
            yshift={-\pgfkeysvalueof{/pgfplots/major tick length} + 4pt},
        },
    }
    \begin{axis}[bar shift=-18pt, nodes near coords style={xshift=-18pt},
        legend pos = outer north east, legend style = {name = pairs}]
        \addplot [fill=force-cpu, nodes near coords=A] table [x=arch, y=pairsfn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=pairsfn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=pairsfn] {data/single-core-other.csv};
        \legend{Force, Neigh, Other}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{A:} P4IRS (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{B:} MD-Bench (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{C:} P4IRS (HN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{D:} MD-Bench (HN)}
    \end{axis}
    \begin{axis}[bar shift= -6pt, nodes near coords style={xshift=-6pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-cpu, nodes near coords=B] table [x=arch, y=mdbenchfn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=mdbenchfn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=mdbenchfn] {data/single-core-other.csv};
    \end{axis}
    \begin{axis}[bar shift= 6pt, nodes near coords style={xshift=6pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-cpu, nodes near coords=C] table [x=arch, y=pairshn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=pairshn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=pairshn] {data/single-core-other.csv};
    \end{axis}
    \begin{axis}[bar shift= 18pt, nodes near coords style={xshift=18pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-cpu, nodes near coords=D] table [x=arch, y=mdbenchhn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=mdbenchhn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=mdbenchhn] {data/single-core-other.csv};
    \end{axis}
\end{tikzpicture}
\vspace{-3ex}
	\caption{Execution time for P4IRS and MD-Bench simulations in seconds (lower is better) of a Copper FCC lattice system with $32^{3}$ unit-cells (4 atoms per unit-cell) during 200 time-steps on Ice Lake and Milan CPUs. Tests were performed in a single core (no parallelism) with fixed frequency. Results are shown for both full and half neighbor-lists (FN/HN).}
\vspace{-2ex}
\centering
\label{fig:md_single_core_runtimes}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.75]
    \pgfplotsset{
        ybar stacked, ymin=0, ymax=1.5, xmin=0.5, xmax=2.5, xtick=data,
        xtick={1,...,2},
        xticklabels={A40, A100},
        xticklabel style={yshift=-10pt},
        ylabel={time to solution (s)}, ylabel style={yshift=-1ex},
        legend cell align={left},
        /pgf/bar width=12pt,% bar width
        scatter/position=absolute,
        node near coords style={
            font=\footnotesize,
            at={(axis cs:\pgfkeysvalueof{/data point/x},\pgfkeysvalueof{/pgfplots/ymin})},
            anchor=north,
            yshift={-\pgfkeysvalueof{/pgfplots/major tick length} + 4pt},
        },
    }
    \begin{axis}[bar shift=-12pt, nodes near coords style={xshift=-12pt},
        legend pos = outer north east, legend style = {name = pairs}]
        \addplot [fill=force-gpu, nodes near coords=A] table [x=arch, y=pairsfn] {data/single-gpu-force.csv};
        \addplot [fill=neigh-gpu] table [x=arch, y=pairsfn] {data/single-gpu-neigh.csv};
        \addplot [fill=other-gpu] table [x=arch, y=pairsfn] {data/single-gpu-other.csv};
        \legend{Force, Neigh, Other}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{A:} P4IRS (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{B:} MD-Bench (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{C:} P4IRS (HN)}
    \end{axis}
    \begin{axis}[bar shift= -0pt, nodes near coords style={xshift=-0pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-gpu, nodes near coords=B] table [x=arch, y=mdbenchfn] {data/single-gpu-force.csv};
        \addplot [fill=neigh-gpu] table [x=arch, y=mdbenchfn] {data/single-gpu-neigh.csv};
        \addplot [fill=other-gpu] table [x=arch, y=mdbenchfn] {data/single-gpu-other.csv};
    \end{axis}
    \begin{axis}[bar shift= 12pt, nodes near coords style={xshift=12pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-gpu, nodes near coords=C] table [x=arch, y=pairshn] {data/single-gpu-force.csv};
        \addplot [fill=neigh-gpu] table [x=arch, y=pairshn] {data/single-gpu-neigh.csv};
        \addplot [fill=other-gpu] table [x=arch, y=pairshn] {data/single-gpu-other.csv};
    \end{axis}
\end{tikzpicture}
\vspace{-3ex}
	\caption{Execution time for P4IRS and MD-Bench simulations in seconds (lower is better) of a Copper FCC lattice system with $50^{3}$ unit-cells (4 atoms per unit-cell) during 200 time-steps on A40 and A100 GPUs. Results are shown for both full and half neighbor-lists (FN/HN) cases, and atomic operations affect the performance for the HN case.}
\vspace{-2ex}
\centering
\label{fig:md_single_gpu_runtimes}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.95]
    \tikzstyle{every node}=[font=\small]
    \begin{axis}[
            width=0.5\textwidth, height=8cm,
            xmode=log,log basis x={2},
            %xmin=0.75,   xmax=3072,
            xmin=1,   xmax=64,
            ymin=10,   ymax=30,
            log ticks with fixed point,
            xlabel={\# nodes}, xlabel style={yshift= 1ex},
            xticklabel={ % hack for precision issues with 1024 and 2048
                \pgfkeys{/pgf/fpu=true}
                \pgfmathparse{int(2^\tick)}
                \pgfmathprintnumber[fixed]{\pgfmathresult}
            },
            ylabel={time to solution (s)}, ylabel style={yshift=-1ex},
            grid=major,
            legend pos=north east,
            %legend style={at={(0.97,0.5)},anchor=east},
        ]
        \addplot table[x=nodes,y=pairsfn] {data/md-scaling-fritz.csv};
        \addplot table[x=nodes,y=pairshn] {data/md-scaling-fritz.csv};
        \legend{P4IRS-FN, P4IRS-HN}
    \end{axis}
\end{tikzpicture}
\vspace{-3ex}
\caption{Weak-scaling results for P4IRS (FN and HN) on Fritz, each node simulates 200 time-steps of a Copper FCC lattice system with $128^3$ unit cells, with approximately 116.508 atoms per CPU core.}
\vspace{-2ex}
\label{fig:md_weak_scaling_fritz}
\end{figure}

@@

\begin{tikzpicture}[scale=0.75]

>>>>

When computing forces for DEM simulations that contain different shapes, different calculation formulas are needed for some of the contact elements such as the penetration depth, contact point and distance.
From a computational point of view, this introduces branching in the inner-most loop where each path leads to different arithmetic instructions be executed and therefore inhibits SIMD vectorization.
Algorithm \ref{alg:branching} displays an example of such issue for a kernel with spheres and half-spaces.
Note that when traversing the neighbors, their shapes are checked to determine which code path is taken during the interaction, hence the code is not SIMD-friendly.

@@

To provide a simpler way to achieve vectorized code for such DEM kernels, we partition particles by shape in our acceleration data structures (Linked Cells and Verlet Lists), which allows distinguishing them during the force computation.
After, we use code generation to apply a loop-splitting in the generated code for every pair of shapes, these resulting loops should be then easier to be vectorized since no branches are needed to calculate different contact elements for the particles being computed.
Algorithm \ref{alg:no_branching} displays the same kernel as Algorithm \ref{alg:branching} without the branching issue.
Note that the outer-most loop can also be split into two parts as well, which leads to two different kernels for each pair of shapes.
For GPUs, this can be beneficial since different threads will be used for different pair of shapes (parallelism is done at the level of the outer-most loop) and no threads will be disabled due to traversing a different code path.

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.73]
    \pgfplotsset{
        ybar stacked, ymin=0, ymax=32, xmin=0.5, xmax=2.5, xtick=data,
        xtick={1,...,2},
        xticklabels={Ice Lake, Milan},
        xticklabel style={yshift=-10pt},
        ylabel={time to solution (s)}, ylabel style={yshift=-1ex},
        legend cell align={left},
        /pgf/bar width=12pt,% bar width
        scatter/position=absolute,
        node near coords style={
            font=\footnotesize,
            at={(axis cs:\pgfkeysvalueof{/data point/x},\pgfkeysvalueof{/pgfplots/ymin})},
            anchor=north,
            yshift={-\pgfkeysvalueof{/pgfplots/major tick length} + 4pt},
        },
    }
    \begin{axis}[bar shift=-18pt, nodes near coords style={xshift=-18pt},
        legend pos = outer north east, legend style = {name = pairs}]
        \addplot [fill=force-cpu, nodes near coords=A] table [x=arch, y=pairsfn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=pairsfn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=pairsfn] {data/single-core-other.csv};
        \legend{Force, Neigh, Other}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{A:} P4IRS (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{B:} MD-Bench (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{C:} P4IRS (HN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{D:} MD-Bench (HN)}
    \end{axis}
    \begin{axis}[bar shift= -6pt, nodes near coords style={xshift=-6pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-cpu, nodes near coords=B] table [x=arch, y=mdbenchfn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=mdbenchfn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=mdbenchfn] {data/single-core-other.csv};
    \end{axis}
    \begin{axis}[bar shift= 6pt, nodes near coords style={xshift=6pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-cpu, nodes near coords=C] table [x=arch, y=pairshn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=pairshn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=pairshn] {data/single-core-other.csv};
    \end{axis}
    \begin{axis}[bar shift= 18pt, nodes near coords style={xshift=18pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-cpu, nodes near coords=D] table [x=arch, y=mdbenchhn] {data/single-core-force.csv};
        \addplot [fill=neigh-cpu] table [x=arch, y=mdbenchhn] {data/single-core-neigh.csv};
        \addplot [fill=other-cpu] table [x=arch, y=mdbenchhn] {data/single-core-other.csv};
    \end{axis}
\end{tikzpicture}
\vspace{-3ex}
	\caption{Execution time for P4IRS and MD-Bench simulations in seconds (lower is better) of a Copper FCC lattice system with $32^{3}$ unit-cells (4 atoms per unit-cell) during 200 time-steps on Ice Lake and Milan CPUs. Tests were performed in a single core (no parallelism) with fixed frequency. Results are shown for both full and half neighbor-lists (FN/HN).}
\vspace{-2ex}
\centering
\label{fig:md_single_core_runtimes}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.73]
    \pgfplotsset{
        ybar stacked, ymin=0, ymax=1.5, xmin=0.5, xmax=2.5, xtick=data,
        xtick={1,...,2},
        xticklabels={A40, A100},
        xticklabel style={yshift=-10pt},
        ylabel={time to solution (s)}, ylabel style={yshift=-1ex},
        legend cell align={left},
        /pgf/bar width=12pt,% bar width
        scatter/position=absolute,
        node near coords style={
            font=\footnotesize,
            at={(axis cs:\pgfkeysvalueof{/data point/x},\pgfkeysvalueof{/pgfplots/ymin})},
            anchor=north,
            yshift={-\pgfkeysvalueof{/pgfplots/major tick length} + 4pt},
        },
    }
    \begin{axis}[bar shift=-12pt, nodes near coords style={xshift=-12pt},
        legend pos = outer north east, legend style = {name = pairs}]
        \addplot [fill=force-gpu, nodes near coords=A] table [x=arch, y=pairsfn] {data/single-gpu-force.csv};
        \addplot [fill=neigh-gpu] table [x=arch, y=pairsfn] {data/single-gpu-neigh.csv};
        \addplot [fill=other-gpu] table [x=arch, y=pairsfn] {data/single-gpu-other.csv};
        \legend{Force, Neigh, Other}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{A:} P4IRS (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{B:} MD-Bench (FN)}
        \addlegendimage{empty legend}
        \addlegendentry{\hspace{-.325cm}\textbf{C:} P4IRS (HN)}
    \end{axis}
    \begin{axis}[bar shift= -0pt, nodes near coords style={xshift=-0pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-gpu, nodes near coords=B] table [x=arch, y=mdbenchfn] {data/single-gpu-force.csv};
        \addplot [fill=neigh-gpu] table [x=arch, y=mdbenchfn] {data/single-gpu-neigh.csv};
        \addplot [fill=other-gpu] table [x=arch, y=mdbenchfn] {data/single-gpu-other.csv};
    \end{axis}
    \begin{axis}[bar shift= 12pt, nodes near coords style={xshift=12pt},
        legend style = {at = {([yshift = -1mm]pairs.south west)},
        anchor = north west}]
        \addplot [fill=force-gpu, nodes near coords=C] table [x=arch, y=pairshn] {data/single-gpu-force.csv};
        \addplot [fill=neigh-gpu] table [x=arch, y=pairshn] {data/single-gpu-neigh.csv};
        \addplot [fill=other-gpu] table [x=arch, y=pairshn] {data/single-gpu-other.csv};
    \end{axis}
\end{tikzpicture}
\vspace{-3ex}
	\caption{Execution time for P4IRS and MD-Bench simulations in seconds (lower is better) of a Copper FCC lattice system with $50^{3}$ unit-cells (4 atoms per unit-cell) during 200 time-steps on A40 and A100 GPUs. Results are shown for both full and half neighbor-lists (FN/HN) cases, and atomic operations affect the performance for the HN case.}
\vspace{-2ex}
\centering
\label{fig:md_single_gpu_runtimes}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.95]
    \tikzstyle{every node}=[font=\small]
    \begin{axis}[
            width=0.5\textwidth, height=8cm,
            xmode=log,log basis x={2},
            %xmin=0.75,   xmax=3072,
            xmin=1,   xmax=64,
            ymin=10,   ymax=30,
            log ticks with fixed point,
            xlabel={\# nodes}, xlabel style={yshift= 1ex},
            xticklabel={ % hack for precision issues with 1024 and 2048
                \pgfkeys{/pgf/fpu=true}
                \pgfmathparse{int(2^\tick)}
                \pgfmathprintnumber[fixed]{\pgfmathresult}
            },
            ylabel={time to solution (s)}, ylabel style={yshift=-1ex},
            grid=major,
            legend pos=north east,
            %legend style={at={(0.97,0.5)},anchor=east},
        ]
        \addplot table[x=nodes,y=pairsfn] {data/md-scaling-fritz.csv};
        \addplot table[x=nodes,y=pairshn] {data/md-scaling-fritz.csv};
        \legend{P4IRS-FN, P4IRS-HN}
    \end{axis}
\end{tikzpicture}
\vspace{-2.5ex}
\caption{Weak-scaling results for P4IRS (FN and HN) on Fritz, each node simulates 200 time-steps of a Copper FCC lattice system with $128^3$ unit cells, with approximately 116.508 atoms per CPU core.}
\vspace{-2ex}
\label{fig:md_weak_scaling_fritz}
\end{figure}

@@

\begin{tikzpicture}[scale=0.72]
